Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-20 treestats_paper.tex
Get rid of empty section to stop Alg B splitting over page.
updated relate disc

The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to the framework described by \citet{fu1995statistical}.
To connect our  generalization to familiar ideas,
here is a reframing of \citeauthor{fu1995statistical}'s statistics.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

@@

In this view,
site statistics are noisy approximations to the corresponding branch statistic
--- but, how noisy?
How big is the contribution of mutation to the overall sampling variance of a statistic?
The law of total variance partitions the variance of a site statistic
into the contributions of noise from demography and mutation:
\begin{align*}
    \var[\site(f, \iw)_{[i,j)}]
    &=
        \var\left[ \E\left[\site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)}\right] \right]
            +
        \E\left[ \var\left[\site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)}\right] \right]
        \\
    &=
        \var\left[ \branch(f, \iw)_{[i,j)} \right]
            +
        \frac{1}{j-i} \E\left[ \branch(f^2, \iw)_{[i,j)} \right] .
\end{align*}
The first term is the variance of the expected Site statistic given the trees,
which by duality is the variance of the Branch statistic,
i.e., the contribution of demographic noise, including genetic drift.
The second term, the expected variance of the statistic given the trees,
is the contribution to variance of mutation,
and can itself be written as a Branch statistic,
using the same weights and the summary function $f(x)^2$
\citep[Lemma 2 in][]{ralph2019empirical}.
The reason for this is straightforward ---
the Site statistic is a sum across branches of the number of mutations on the branch
multiplied by a summary, $f(x)$.
Given the trees, these numbers of mutations are independent and Poisson,
and the variance of $f(x)$ multiplied by a Poisson random variable with mean $m$ is $m f(x)^2$.
The sum is divided by the length of the region, $j-i$,
and so the variance is divided by $(j-i)^2$.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that Branch and Site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between Site and Branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diversity using the Site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual Branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method
(which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

%%%%%%%%%%%%%%%%%%
\section*{Efficient algorithms and implementation}

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyze the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across Site, Branch and Node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

>>>>

The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to the linear functions of the frequency spectrum
discussed by \citet{fu1995statistical} and \citet{achaz2009frequency}.
To connect our generalization to familiar ideas,
here is a reframing of these statistics of the frequency spectrum.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

@@

In this view,
site statistics are noisy approximations to the corresponding branch statistic
--- but, how noisy?
How big is the contribution of mutation to the overall sampling variance of a statistic?
The law of total variance partitions the variance of a site statistic
into the contributions of noise from demography and mutation:
\begin{align*}
    \var[\site(f, \iw)_{[i,j)}]
    &=
        \var\left[ \E\left[\site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)}\right] \right]
            +
        \E\left[ \var\left[\site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)}\right] \right]
        \\
    &=
        \var\left[ \branch(f, \iw)_{[i,j)} \right]
            +
        \frac{1}{j-i} \E\left[ \branch(f^2, \iw)_{[i,j)} \right] .
\end{align*}
The first term is the variance of the expected Site statistic given the trees,
which by duality is the variance of the Branch statistic,
i.e., the contribution of demographic noise, including genetic drift.
The second term, the expected variance of the statistic given the trees,
is the contribution to variance of mutation,
and can itself be written as a Branch statistic,
using the same weights and the summary function $f(x)^2$
\citep[Lemma 2 in][]{ralph2019empirical}.
The reason for this is straightforward ---
the Site statistic is a sum across branches of the number of mutations on the branch
multiplied by a summary, $f(x)$.
Given the trees, these numbers of mutations are independent and Poisson,
and the variance of the product of the constant $f(x)$
and a Poisson random variable with mean $m$ is $m f(x)^2$.
The sum is divided by the length of the region, $j-i$,
and so the variance is divided by $(j-i)^2$.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that Branch and Site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between Site and Branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diversity using the Site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual Branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, the ratio of Site and Branch diversity
is relatively constant, hovering around typical estimates for the average genome-wide human mutation rate~\citep{narasimhan2017estimating}.
On the one hand, this is expected ---
Relate estimates branch lengths in part assuming that mutations fall at a constant rate through time
(although \citeauthor{speidel2019method} also showed signal of changing mutation rates,
pointing the way towards a more general method).
On the other hand,
since estimated node ages are shared across long regions of the genome,
a tight agreement may not be possible with poorly inferred trees.
So, the relative constancy of the ratio of Site to Branch diversity
suggests that Relate is doing a good job at inferring trees,
but what to make of the twofold variation in this ratio?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

%%%%%%%%%%%%%%%%%%
\section*{Algorithm and implementation}

In this section we describe and analyze the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across Site, Branch and Node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.
